A Classical Analog of Quantum Interaction-Free Measurement

### 2015-07-17 ms/ifm.tex
editing

There are also counter-intuitive quantum phenonema not involving
entanglement.  Imagine a photon, initially in a ``vertical'' polarized
state.  It is sent through a Faraday rotator, and then through a
vertically polarizing sheet.  The former rotates the polarization
plane towards ``horizontal'' by a small angle, say $\pi/(2N)$, where
$N$ is large.  The latter effectively measures the polarization state
in the vertical/horizontal basis; if the result comes out horizontal,
the photon is absorbed, otherwise it goes through in a vertical
eigenstate.  Since $N$ is large, the chances of being absorbed are
small.  More specifically, after the rotation the {\em amplitude\/} to
be in the horizontal state will be $\sim1/N$ so the probability to be
in that mode --- and therefore the probability of absorption --- will
be $\sim1/N^2$.  Thus, most likely, the photon will get slightly
rotated and then reset to vertical.  If the photon is re-routed back
to the apparatus $N$ times, it will still most likely emerge
vertically polarized.  The rotator and the polarizer appear to cancel
each other out.  The general phenomenon is known as the quantum Zeno
effect,\cite{PhysRevA.41.2295} but it can also be described in
everyday language, in terms of waking and sleeping
states.\cite{orzel2009teach}

@@

In this paper, we argue that interaction-free measurement need not be
quantum at all.  Rather, the characteristics features of (a)~an
amplitude that cycles between two states, but (b)~an absorber can
block this cycling, thus revealing its presence, but without actually
absorbing any energy, can also appear in classical systems.  If one
has a pair of identical coupled oscillators $x$ and $y$, and $x$ is
set in motion, one will see the strength of the oscillation of $x$
slowly decrease with time, and over longer times the energy will
periodically `slosh' back and forth between the two systems.  But if
instead oscillator $y$ is damped --- so it is able to {\em absorb\/}
energy --- the rate of decrease of the energy in oscillator $x$ will
be suppressed.  Thus one can infer the ability of oscillator $y$ to
absorb energy by observing the behaviour of oscillator $x$ without, in
the limit that the damping is strong, there actually being any energy
absorbed.

The above thought experiment involving rotation the polarization plane
of photon is one simple way to approach the concept of
interaction-free measurement, but not the only one.  Another
well-known description, which more closely resembles some of the
actual experiments, is in terms of a Mach-Zehnder interferometer.  The
horizontal and vertical polarization states are replaced by the two
arms of the interferometer.  Figure~\ref{fig:mz} schematically shows
the $N=1$ case; the general case would be a concatenation of such
devices.

\section{An interferometer as an absorber detector}

@@

\section{Coupled oscillators as an absorber detector}

>>>>

There are also counter-intuitive quantum phenonema not involving
entanglement.  Imagine a photon, initially in a ``vertical'' polarized
state.  It is sent through a Faraday rotator, and then through a
vertically polarizing sheet.  The former rotates the polarization
plane towards ``horizontal'' by a small angle, say $\pi/(2N)$, where
$N$ is large.  The latter effectively measures the polarization state
in the vertical/horizontal basis; if the result comes out horizontal,
the photon is absorbed, otherwise it goes through in a vertical
eigenstate.  Since $N$ is large, the chances of being absorbed are
small.  More specifically, after the rotation the {\em amplitude\/} to
be in the horizontal state will be $\sim1/N$ so the probability to be
in that mode --- and therefore the probability of absorption --- will
be $\sim1/N^2$.  Thus, most likely, the photon will get slightly
rotated and then reset to vertical.  If the photon is re-routed back
to the apparatus $N$ times, it will still most likely emerge
vertically polarized.  The rotator and the polarizer appear to cancel
each other out.  The general phenomenon is known as the quantum Zeno
effect,\cite{1977JMP....18..756M,PhysRevA.41.2295} but it can also be
described in everyday language, in terms of waking and sleeping
states.\cite{orzel2009teach}

@@

In this paper, we argue that interaction-free measurement need not be
quantum at all.  Rather, the characteristics features of (a)~an
amplitude that cycles between two states, but (b)~an absorber can
block this cycling, thus revealing its presence, but without actually
absorbing any energy, can also appear in classical systems.  If one
has a pair of identical coupled oscillators $x$ and $y$, and $x$ is
set in motion, one will see the strength of the oscillation of $x$
slowly decrease with time, and over longer times the energy will
periodically slosh back and forth between the two systems.  But if
instead oscillator $y$ is damped --- so it is able to {\em absorb\/}
energy --- the rate of decrease of the energy in oscillator $x$ will
be suppressed.  Thus one can infer the ability of oscillator $y$ to
absorb energy by observing the behaviour of oscillator $x$ without, in
the limit that the damping is strong, there actually being any energy
absorbed.

In the other sections of this paper, we flesh out the above argument.
In section~\ref{sec:interf} we work through an alternative well-known
description of interaction-free measurement, more closely resembles
some of the actual experiments.  This is in terms of a Mach-Zehnder
interferometer, whose two arms replace the horizontal and vertical
polarization states of the above thought-experiment.
Figure~\ref{fig:mz} schematically shows the $N=1$ case; the general
case would be a concatenation of such devices.  With an absorber, the
amplitude sloshes between the 'lower' and 'upper' arms of the
interferometer; the absorber results in confining the amplitude to the
lower arm.  Section~\ref{sec:oscill} briefly solves for the
oscillators $x$ and $y$ from the previous paragraph.

\section{An interferometer as an absorber detector}\label{sec:interf}

@@

\section{Coupled oscillators as an absorber detector}\label{sec:oscill}
